Overview and Status of Assumptions of Physics\footnote{http://assumptionsofphysics.org} \\ \ifextended (Extended version) \else (Short version) \fi

### 2019-12-15 Articles/Overview/overview.tex
Double version for overview

\title{Overview and Status of the Assumptions of Physics}
\author{Gabriele Carcassi, Christine A. Aidala}
\date{November 2019}

@@

\emph{Not every subject can be studied by science.} The principle indirectly states that anything that is subjective, contradictory or not evidence based cannot be the subject of scientific investigation. The properties of prime numbers, moral and existential questions or the aesthetics of music fail in at least one of those respects and are excluded. Only the parts of nature that are accessible through consistent experimental verification can be studied scientifically. New experimental techniques have been extending that reach over time.

@@

\emph{Note on terminology.} The terminology may depart slightly from what logicians may be accustomed to, simply because our aim is different. Namely, our goal is not to study the rule of inference or study what can be formally proven, but simply to keep track of the logical relationships between statements that are assumed to be already given.

\emph{Statements as primary object.} Every formal theory needs prime objects, elements that will not further be described by the theory. We will take statements, assertion that are either true or false for everybody, as these objects. The formal system will specifically:
\begin{description}
	\item not be propositional; the language of the statements, its syntax, its grammar, will be left unspecified; statements will represent the content regardless of how it is expressed;
	\item be algebraic; operations on statements will not form new statements, but describe relationships within a given set of statements; note that most mathematical structures used in science are algebraic (computer science is a notable exception)
	\item purely formal; the semantic of the statements is not captured, though the semantic will impose logical relationships that will be formally captured
\end{description}

\emph{Statements cannot exist in isolation.} Statements should, in general, not be regarded as independent of the context they are part of. Take the statement ``The mass of the electron is 510 $\pm$ 0.5 KeV''. If we are measuring the mass of the electron, that is a statement that may or may not be true depending on experiment. If we are performing particle identification in a detector, that statement is assumed to be true. The truth values allowed by the same statement depends on context.

\emph{Logical context as the fundamental structure.} A logical context is a set of statements with well defined logical relationships. It defines which truth assignments are possible (i.e. are consistent with the semantic of the statements). It identifies one of the possible assignments as the true one. It allows to find statements whose truth value depends on others through a particular Boolean function. The existence of this structure is introduced by axioms as they are crossing the lines between physical objects and mathematical ones. We currently do not assume there is a single universal context.

\emph{Tautologies, contradictions and contingent statements.} We define tautologies as those statements that are true in every assignment, contradictions as those statements that are false in every assignment and contingent all the remaining.

\emph{Statement equivalence and Boolean algebra.} Within our formal system, we can distinguish between:
\begin{description}
	\item statement equality: two statements that re-express the same fact in different languages or in different words
	\item statement equivalence: two statements have the same truth value in all possible assignments
	\item statement material equivalence: two statements have the same truth value
\end{description}
A logical context is a complete Boolean algebra under statement equivalence.

\emph{Narrowness and ordering.} A statement is narrower then another if whenever it is true then the other is also true. Narrowness can often be thought as implication, except in a few corner case (e.g. a statement that is always false is narrower than any other statement). Narrowness imposes a partial order to the structure. This is the same order associated to the fact that every Boolean algebra is a complemented distributed lattice.

\emph{Compatibility and independence.} Two statements are said to be compatible if they can both be true in the same assignment. Two statements are independent if fixing the truth value of one does not restrict the truth value of the other. Independence, like linear independence or statistical independence of which it is the foundation, is a property of a set of statements and is not transitive.

@@

\subsubsection{Experimental domains}

@@

\emph{Tests for the relationship domain.} The experimental verifiability of relationship domain statements cannot be proven from the verifiability of each domain. This formalizes the problem that being able to measure a single point (i.e. the temperature and height in a particular circumstance) is not enough to verify the relationship (i.e. the temperature and height are always linked in a particular way). The ability to experimentally explore the space is what builds confidence, which may not be always possible. What constitutes enough evidence is left, as before, outside of the formalism.

@@

This part has not yet been formally developed, though some common ideas are starting to emerge.

\emph{All states are equilibrium states.} In thermodynamics equilibria are simply assumed to exist, without specifying exactly how they are achieved. However, in all physical theories, states are assumed to exist, without specifying what are the conditions upon which the system can be characterized by such states. The idea that, at the very least, all states must be equilibrium states of processes that happen at a faster scale than the processes than we are considering.\footnote{For example, the position and velocity of a book makes sense on the surface of the earth, at a temperature of 293 Kelvin. On the surface of the sun, we would not be able to talk about the position and velocity of a book.} This sets the scale of infinitesimal time intervals.

\emph{States are objective but not absolute.} The same system can be found and studied in different conditions. Depending on the circumstances we may be interested in studying different properties of the system.\footnote{Consider a ball. If we are studying its motion, the state can be typically be captured by position and momentum of the center of mass. If the ball is rotating at high speeds in air, the rotation speed and axis will also need to be part of the state because of the Magnus effect. If we are studying the gas within the ball, the state may be given by volume, pressure and temperature. If the ball is so tiny and light that random impact of the fluids on the surface will affect the motion, the state will be a statistical ensemble. And so on.} Therefore the state will change depending on the situation and our focus, and in this sense it is not absolute. However, once the situation and focus are well defined, the state is objective in the sense that everybody will come to the same characterization.

\emph{The tension between isolation and interaction.} Properties of a system can be fully characterized if they are not perturbed by the interaction with the environment, if they are in stable equilibrium.\footnote{For example, the position of the center of mass of a cannonball is not affected by the air and can therefore be characterized by a set of numbers; however, the position of the center of mass of a spek of dust is affected by the air and must be characterized by a probability distribution over the possible values.} To fully characterize a system, then, we would want it to be in perfect isolation.\footnote{Note that full isolation is not possible as this would mean no interaction, not even gravitation, between the system and the environment, to the point that the system has essentially vanished.} However, to be able to measure those quantities we would like to interact with the system. To fully access the system, then, we would want it to be in constant interaction with the environment. Conceptually, there is a trade-off. Specifying the state space means characterizing this trade-off for the specific system in the specific circumstances.

@@

\emph{Composite states are distributions over particle states.} Under infinitesimal reducibility, if $\mathcal{C}$ is the state space of the while system and $\mathcal{S}$ is the state space of the particles, then each state $c \in \mathcal{C}$ for the full system is identified by a distribution $\rho : \mathcal{S} \to \mathbb{R}$ over the states of infinitesimal parts. The function is real to signify that we can associate arbitrarily small amounts to each particle states. The distribution tells us how many particles can be found in each particle state.

\emph{Continuous time implies particle state space is a manifold.} If time is assumed to be continuous, then the state space $\mathcal{S}$ is a manifold. That is, each particle state $s \in U \subseteq \mathcal{S}$ can be identified by a set of continuous quantities $\xi^a : U \to \mathbb{R}$ which we call state variables. It is a consequence of the general theory that, if we wanted to write a trajectory $\xi^a(t)$, then this must be a topologically continuous function. Therefore, conceptually, once we assume time to be a real number, the state space of the particles must be charted by a set of real quantities: the topology that is given to time reasserts itself on all the quantities that will depend on time.

\emph{Densities imply differentiability.} As we can write the state $s(\xi^a)$ as a function of the state variables $\xi^a$, we will also want to write the density $\rho(s(\xi^a)) = \rho(\xi^a)$ as a function of the state variables. This requires the state variables to be differentiable with respect to the distribution and with each other. Recall that if the Jacobian is not well defined during a change of variable $\rho(\hat{\xi}^b) = \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a}\right| \rho(\xi^a)$ then the density wouldn't be expressible in the new coordinates. Mathematically, this means that the manifold is equipped with a differentiable structure. Note that any time evolution that would map a density to another density will not only have to be continuous, but also differentiable. In short: assuming density distributions as our primary objects, instead of points, justifies why differentiable structures and differentiable time evolution is present in physics.

\emph{Coordinate invariant densities imply phase-space.} As particles states must be defined independently of coordinate systems, they must be invariant under coordinate transformations. That is, $s(\xi^a) = s(\hat{\xi}^b(\xi^a))$. As the density distribution depends only on the state, it should be invariant as well, that is $\rho(s(\xi^a)) = \rho(s(\hat{\xi}^b(\xi^a)))$. We have two seemingly contradicting requirement: $\rho$ should both change like a density and be invariant. The only way to satisfy both requirements is the following: particle states are fully identified by a set of coordinates $q^i$ and by conjugate set of variables $k_i$ whose units are the inverse of the ones defined by $q^i$. This way the product $dq^i dk_i$ is a pure number that is invariant under coordinate transformations. A coordinate transformation $\hat{q}^j = \hat{q}^j (q^i)$ induces a transformation $\hat{k}_j = \frac{\partial q^i }{\partial \hat{q}^j} k_i$ such that the area element $d\hat{q}^j d\hat{k}_j = dq^i \frac{\partial \hat{q}^j }{\partial q^i} \frac{\partial q^i }{\partial \hat{q}^j} dk_i=dq^i dk_i$ is invariant. An independent degree of freedom, then, is a pair $(q,k)$ for which the units can be changed independently from all other state variables. Mathematically, symplectic manifolds are the only spaces that allow us to define coordinate invariant densities. We can introduce a constant $\hbar$ such that $p_i = \hbar k_i$ which at this point is simply an arbitrary choice of unit.

These results gives us a precise understanding of what each layer of the mathematical structure corresponds to physically:
\begin{description}[noitemsep]
	\item \emph{Set of points:} possible states for the particles.
	\item \emph{Topological space:} ability to experimentally distinguish the states.
	\item \emph{Manifold:} states can be distinguished with a set of real valued quantities or state variables.
	\item \emph{Differentiable manifold:} we can define a distribution over the states and the state variables.
	\item \emph{Symplectic manifold:} we can define a distribution over the states that are coordinate independent.
\end{description}

@@

\emph{Reversibility as reconstructing the past.} This definition of reversibility should not be confused with time symmetry or the ability to undo the change. The first one means that if one substitute $t \to -t$ the dynamics is invariant (i.e. the same) or form invariant (i.e. different but described by the same equations) depending on the author. The second one means that we can find another process that brings the final state back to the original state. These concepts do not seem to work well in the general case. For example, when discussing reversibility in the sense of undoing the changes, one typically discusses changing the direction of the velocities to bring about the initial state with inverted velocities. In the presence of forces or an arbitrary dynamics, this operation because more complicated or ill-defined. The core problem with this definition is that it is not a property of the process itself, but on the existence of other processes (i.e. the process that undoes the changes or the process that flips the velocities, therefore changing the state). The requirement of time-invariance is too restrictive as it excludes all time dependent processes. Form invariance on an arbitrary set of equations, instead, can be broken by a suitably complicated change of variables, so it is only well-defined for a limited number of systems (e.g. particles under potential forces). The definition of reversibility we use is preferable because it works more generally and, when appropriate conditions apply, reduces to the other two which are better regarded as special cases. It is mathematically more elegant as it is the dual of determinism. It connects directly with information theory, as it states that the present has enough information to reconstruct the past, which gives a logical bridge to connect reversibility with information entropy and thermodynamics/statistical mechanics entropy.

\emph{Determinism and reversibility imply Hamiltonian mechanics.} In the context of infinitesimal reducibility, determinism and reversibility means that all the particles and only the particle within a given initial state $\rho(s(t)) = \rho(s(t+\Delta t))$ will be mapped to a particular final state. That is, the densities will be preserved over time. Time evolution, then, is a canonical transformation in phase space. As time is continuous, we can find a function $H : \mathcal{S} \to \mathbb{R}$ which acts as a time generator, which we recognize as being the Hamiltonian. This is essentially Louisville's theorem applied in reverse.

\emph{Invariant densities under time transformations imply extended phase-space.} So far, time has been treated as a parameter. If we introduce coordinate transformation that mix time, however, this is no longer possible. If $d\hat{q} = dq + vdt$ then the densities need to be defined on $dt$ as well. To do that, we have to introduce a conjugate variable $\omega$ of unit inverse time so that $dt d\omega$ is dimensionless and invariant. The invariant area element becomes $\sum_i dq^i dk_i - dt d\omega$ or $\sum_i dq^i dp_i - dt dE$ in units of $\hbar$. For surfaces at equal time $dt=0$ it reduces to the previous case. For surfaces that are not at equal time, since the temporal degree of freedom does cannot introduce new states, we need to correct the area element to account for the larger areas measured by the state variables that does not correspond to new states. If we set $q^\alpha = [t, q^i]$ and $p_\alpha = [-E, p_i]$ the area element becomes $dq^\alpha dp_\alpha$. These are elements of special relativity that come just by requiring invariance under time transformation. They do not come from invariance of the speed of light, they do not require a metric tensore. In other words, there is no way to have time transformations in Hamiltonian mechanics without requiring a four-momentum co-vector. Special relativity is not an independent choice: it is already partially baked into the two assumptions we took.

\emph{Extended phase-space imply classical anti-particles.} The trajectories in the extended phase space are in terms of an affine parameter $s$ and not time: $q^\alpha(s)$. As the motion is still deterministic and reversible, though, time $t(s)$ must be monotonic with respect to the affine parameter. For each state, the direction of time can be either aligned or anti-aligned with the affine parameter. Because the relationship is monotonic, aligned states can only connect to other aligned states and anti-aligned states can only connect to other anti-aligned states. We call aligned states the particle states and the anti-aligned states the anti-particle states.

\emph{Time evolution is better expressed by the Hamiltonian constraint.} The temporal degree of freedom introduces two variables but, since one is energy, there is an additional constraint since the value of the energy is linked to the other variables through the Hamiltonian. This link is better expressed by the Hamiltonian constraint $\mathcal{H} : \mathcal{\bar{S}} \to \mathbb{R}$ which is a function over the extended phase space, just like the Hamiltonian was a function over the standard phase space. The Hamiltonian constraint has two roles: it gives us the equation of motions in the same way of the standard Hamiltonian, and it constraints the motion over the surface $\mathcal{H} = 0$. To give context, the Hamiltonian constraint for a free particle is $\mathcal{H} = \frac{|p^\alpha|^2}{2m} - \frac{1}{2}mc^2$, which constraints the norm of the four-momentum to $mc$. The Hamiltonian constraint, unlike the standard Hamiltonian, is invariant under time transformation. It is the generator of the evolution in terms of an affine parameter.

\emph{Hamiltonian constraint as operator on distributions.} If $\rho$ is the distribution for the whole system, then $\mathcal{H} \rho = 0$. In fact, the distribution can only be non-zero on the constrain, for which the Hamiltonian constraint is zero. So, for each point in the extended phase space, either the distribution is zero or the Hamiltonian constrain is zero. Note that this is the same form one has in quantum mechanics for the Klein-Gordon and Dirac equations $\mathcal{H} |\psi\rangle = 0$, which in fact can be understood as Hamiltonian constraints that fix the norm of the momentum four-vector and four-velocity.

\emph{Determinism and reversibility is equivalent to conservation of information entropy.} Given a distribution, its information entropy is given by $I = - \int \rho \log \rho \, d\xi^n$. In general, information entropy is not invariant under change of variable: $- \int \rho(\hat{\xi}^b) \log \rho(\hat{\xi}^b) \, d\hat{\xi}^n = - \int \rho(\xi^a) \log \rho(\xi^a) \, d\xi^n - \int \rho(\xi^i) \log \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a}\right| \, d\xi^n$. Under a canonical transformation over phase space, though, the Jacobian determinant is unitary everyone therefore information entropy is invariant. That is, invariance of the density and invariance of its information entropy are one and the same. Requesting that densities are conserved in time is therefore equivalent to requesting that information entropy is conserved in time. This should be intuitive: the amount of information needed to identify an element within the initial or the final distribution is the same if the evolution is deterministic and reversible. Deterministic and reversible evolution, Hamiltonian evolution and conservation of information entropy, then, are one and the same.

\emph{Conservation of information entropy leads to classical uncertainty principle.} As information entropy is conserved, a Hamiltonian process can only map a distribution to one that has the same information entropy. If we consider a single degree of freedom $(q, p)$ and fix the information entropy $I_0$, then the distribution that minimizes the product of the standard deviation is the Gaussian. The information entropy for the Gaussian is given by $I_G = \ln (2\pi e \sigma_q \sigma_p)$. During the evolution, we have:
$$ \sigma_q \sigma_p \geq \frac{e^{I_0}}{2\pi e} $$.

@@

Note that we use $(q^\alpha, p_\alpha)$ for position and momentum and $(x^\alpha, u^\alpha)$ for position and velocity even though $q^\alpha = x^\alpha$. We do this because $\frac{\partial}{\partial q^\alpha} \neq \frac{\partial}{\partial x^\alpha}$ since one is taken at constant $p_\alpha$ and the other at constant $u^\alpha$. This is a tremendous source of confusion which is avoided by the use of the two symbols.

\emph{Kinematic equivalence means having an invertible relationship between velocity and momentum.} If the state of the system $(q^\alpha, p_\alpha)$ is enough to determine the trajectory $x^\alpha(s)$, then it will be enough to determine the position and velocity $(x^\alpha, u^\alpha)$. These will be enough to reconstruct the state as they span a space of equal dimension. As $q^\alpha = x^\alpha$, then $u^\alpha = u^\alpha(q^\beta, p_\gamma)$ is invertible.

>>>>

\newif\ifextended

\extendedfalse

\newcommand\fact[2]{\ifextended \emph{#1}.#2 \fi}

\title{Overview and Status of Assumptions of Physics\footnote{http://assumptionsofphysics.org} \\
\ifextended (Extended version)
\else (Short version)
\fi}
\author{Gabriele Carcassi, Christine A. Aidala}
\date{November 2019}

@@

\fact{Not every subject can be studied by science} {The principle indirectly states that anything that is subjective, contradictory or not evidence based cannot be the subject of scientific investigation. The properties of prime numbers, moral and existential questions or the aesthetics of music fail in at least one of those respects and are excluded. Only the parts of nature that are accessible through consistent experimental verification can be studied scientifically. New experimental techniques have been extending that reach over time.}

@@

\fact{Note on terminology}{The terminology may depart slightly from what logicians may be accustomed to, simply because our aim is different. Namely, our goal is not to study the rule of inference or study what can be formally proven, but simply to keep track of the logical relationships between statements that are assumed to be already given.}

\fact{Statements as primary object} {Every formal theory needs prime objects, elements that will not further be described by the theory. We will take statements, assertion that are either true or false for everybody, as these objects. The formal system will specifically:
\begin{description}
	\item not be propositional; the language of the statements, its syntax, its grammar, will be left unspecified; statements will represent the content regardless of how it is expressed;
	\item be algebraic; operations on statements will not form new statements, but describe relationships within a given set of statements; note that most mathematical structures used in science are algebraic (computer science is a notable exception)
	\item purely formal; the semantic of the statements is not captured, though the semantic will impose logical relationships that will be formally captured
\end{description}}

\fact{Statements cannot exist in isolation} {Statements should, in general, not be regarded as independent of the context they are part of. Take the statement ``The mass of the electron is 510 $\pm$ 0.5 KeV''. If we are measuring the mass of the electron, that is a statement that may or may not be true depending on experiment. If we are performing particle identification in a detector, that statement is assumed to be true. The truth values allowed by the same statement depends on context.}

\fact{Logical context as the fundamental structure} {A logical context is a set of statements with well defined logical relationships. It defines which truth assignments are possible (i.e. are consistent with the semantic of the statements). It identifies one of the possible assignments as the true one. It allows to find statements whose truth value depends on others through a particular Boolean function. The existence of this structure is introduced by axioms as they are crossing the lines between physical objects and mathematical ones. We currently do not assume there is a single universal context.}

\fact{Tautologies, contradictions and contingent statements} {We define tautologies as those statements that are true in every assignment, contradictions as those statements that are false in every assignment and contingent all the remaining.}

\fact{Statement equivalence and Boolean algebra} {Within our formal system, we can distinguish between:
\begin{description}
	\item statement equality: two statements that re-express the same fact in different languages or in different words
	\item statement equivalence: two statements have the same truth value in all possible assignments
	\item statement material equivalence: two statements have the same truth value
\end{description}
A logical context is a complete Boolean algebra under statement equivalence.}

\fact{Narrowness and ordering} {A statement is narrower then another if whenever it is true then the other is also true. Narrowness can often be thought as implication, except in a few corner case (e.g. a statement that is always false is narrower than any other statement). Narrowness imposes a partial order to the structure. This is the same order associated to the fact that every Boolean algebra is a complemented distributed lattice.}

\fact{Compatibility and independence} {Two statements are said to be compatible if they can both be true in the same assignment. Two statements are independent if fixing the truth value of one does not restrict the truth value of the other. Independence, like linear independence or statistical independence of which it is the foundation, is a property of a set of statements and is not transitive.}

@@

\subsubsection{Experimental and theoretical domains}

@@

\fact{Tests for the relationship domain} {The experimental verifiability of relationship domain statements cannot be proven from the verifiability of each domain. This formalizes the problem that being able to measure a single point (i.e. the temperature and height in a particular circumstance) is not enough to verify the relationship (i.e. the temperature and height are always linked in a particular way). The ability to experimentally explore the space is what builds confidence, which may not be always possible. What constitutes enough evidence is left, as before, outside of the formalism.}

@@

This part has not yet been formally developed, though some common ideas are starting to emerge. The main idea comes from the realization that, when defining a system, the notions of states, processes, time, evolution laws, interaction with the environment and equilibria are all interrelated and can't be given independently. Roughly speaking, to define a system we have to choose a boundary that separates it from the environment. Only the quantities that are unaffected by the interaction with the environment can, in those circumstances, be thought as properties of the system. This in turns is linked to what we can define as state for the system, which is linked to the laws we can write. The processes and the laws are naturally linked to what is assumed about time.

The idea, then, is that giving a set of states is actually giving a much richer structure. It means defining a set of interaction at the boundary, it means defining equilibria with such interactions, it means defining time-scales below which and above which our description fails or is not interesting, it means defining how probing the system may affect those assumptions.\footnote{If we are studying a molecule, we may impose boundaries and processes such that that molecule and its constituent remain stable, which means, for example, interaction energies lower than the chemical bonds.} The plan is, at some point, to formalize these objects and relationships to provide a basic mathematical structure that all physical theories must adhere to.

\fact{All states are equilibrium states} {In thermodynamics equilibria are simply assumed to exist, without specifying exactly how they are achieved. However, in all physical theories, states are assumed to exist, without specifying what are the conditions upon which the system can be characterized by such states. The idea that, at the very least, all states must be equilibrium states of processes that happen at a faster scale than the processes than we are considering.\footnote{For example, the position and velocity of a book makes sense on the surface of the earth, at a temperature of 293 Kelvin. On the surface of the sun, we would not be able to talk about the position and velocity of a book.} This sets the scale of infinitesimal time intervals.}

\fact{States are objective but not absolute} {The same system can be found and studied in different conditions. Depending on the circumstances we may be interested in studying different properties of the system.\footnote{Consider a ball. If we are studying its motion, the state can be typically be captured by position and momentum of the center of mass. If the ball is rotating at high speeds in air, the rotation speed and axis will also need to be part of the state because of the Magnus effect. If we are studying the gas within the ball, the state may be given by volume, pressure and temperature. If the ball is so tiny and light that random impact of the fluids on the surface will affect the motion, the state will be a statistical ensemble. And so on.} Therefore the state will change depending on the situation and our focus, and in this sense it is not absolute. However, once the situation and focus are well defined, the state is objective in the sense that everybody will come to the same characterization.}

\fact{The tension between isolation and interaction.} {Properties of a system can be fully characterized if they are not perturbed by the interaction with the environment, if they are in stable equilibrium.\footnote{For example, the position of the center of mass of a cannonball is not affected by the air and can therefore be characterized by a set of numbers; however, the position of the center of mass of a spek of dust is affected by the air and must be characterized by a probability distribution over the possible values.} To fully characterize a system, then, we would want it to be in perfect isolation.\footnote{Note that full isolation is not possible as this would mean no interaction, not even gravitation, between the system and the environment, to the point that the system has essentially vanished.} However, to be able to measure those quantities we would like to interact with the system. To fully access the system, then, we would want it to be in constant interaction with the environment. Conceptually, there is a trade-off. Specifying the state space means characterizing this trade-off for the specific system in the specific circumstances.}

@@

\fact{Composite states are distributions over particle states} {Under infinitesimal reducibility, if $\mathcal{C}$ is the state space of the while system and $\mathcal{S}$ is the state space of the particles, then each state $c \in \mathcal{C}$ for the full system is identified by a distribution $\rho : \mathcal{S} \to \mathbb{R}$ over the states of infinitesimal parts. The function is real to signify that we can associate arbitrarily small amounts to each particle states. The distribution tells us how many particles can be found in each particle state.}

\fact{Continuous time implies particle state space is a manifold} {If time is assumed to be continuous, then the state space $\mathcal{S}$ is a manifold. That is, each particle state $s \in U \subseteq \mathcal{S}$ can be identified by a set of continuous quantities $\xi^a : U \to \mathbb{R}$ which we call state variables. It is a consequence of the general theory that, if we wanted to write a trajectory $\xi^a(t)$, then this must be a topologically continuous function. Therefore, conceptually, once we assume time to be a real number, the state space of the particles must be charted by a set of real quantities: the topology that is given to time reasserts itself on all the quantities that will depend on time.}

\fact{Densities imply differentiability} {As we can write the state $s(\xi^a)$ as a function of the state variables $\xi^a$, we will also want to write the density $\rho(s(\xi^a)) = \rho(\xi^a)$ as a function of the state variables. This requires the state variables to be differentiable with respect to the distribution and with each other. Recall that if the Jacobian is not well defined during a change of variable $\rho(\hat{\xi}^b) = \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a}\right| \rho(\xi^a)$ then the density wouldn't be expressible in the new coordinates. Mathematically, this means that the manifold is equipped with a differentiable structure. Note that any time evolution that would map a density to another density will not only have to be continuous, but also differentiable. In short: assuming density distributions as our primary objects, instead of points, justifies why differentiable structures and differentiable time evolution is present in physics.}

\fact{Coordinate invariant densities imply phase-space} {As particles states must be defined independently of coordinate systems, they must be invariant under coordinate transformations. That is, $s(\xi^a) = s(\hat{\xi}^b(\xi^a))$. As the density distribution depends only on the state, it should be invariant as well, that is $\rho(s(\xi^a)) = \rho(s(\hat{\xi}^b(\xi^a)))$. We have two seemingly contradicting requirement: $\rho$ should both change like a density and be invariant. The only way to satisfy both requirements is the following: particle states are fully identified by a set of coordinates $q^i$ and by conjugate set of variables $k_i$ whose units are the inverse of the ones defined by $q^i$. This way the product $dq^i dk_i$ is a pure number that is invariant under coordinate transformations. A coordinate transformation $\hat{q}^j = \hat{q}^j (q^i)$ induces a transformation $\hat{k}_j = \frac{\partial q^i }{\partial \hat{q}^j} k_i$ such that the area element $d\hat{q}^j d\hat{k}_j = dq^i \frac{\partial \hat{q}^j }{\partial q^i} \frac{\partial q^i }{\partial \hat{q}^j} dk_i=dq^i dk_i$ is invariant. An independent degree of freedom, then, is a pair $(q,k)$ for which the units can be changed independently from all other state variables. Mathematically, symplectic manifolds are the only spaces that allow us to define coordinate invariant densities. We can introduce a constant $\hbar$ such that $p_i = \hbar k_i$ which at this point is simply an arbitrary choice of unit.}

\fact{Precise physics-math correspondence} {These results gives us a precise understanding of what each layer of the mathematical structure corresponds to physically:
\begin{description}[noitemsep]
	\item \emph{Set of points:} possible states for the particles.
	\item \emph{Topological space:} ability to experimentally distinguish the states.
	\item \emph{Manifold:} states can be distinguished with a set of real valued quantities or state variables.
	\item \emph{Differentiable manifold:} we can define a distribution over the states and the state variables.
	\item \emph{Symplectic manifold:} we can define a distribution over the states that are coordinate independent.
\end{description}}

@@

\fact{Reversibility as reconstructing the past} {This definition of reversibility should not be confused with time symmetry or the ability to undo the change. The first one means that if one substitute $t \to -t$ the dynamics is invariant (i.e. the same) or form invariant (i.e. different but described by the same equations) depending on the author. The second one means that we can find another process that brings the final state back to the original state. These concepts do not seem to work well in the general case. For example, when discussing reversibility in the sense of undoing the changes, one typically discusses changing the direction of the velocities to bring about the initial state with inverted velocities. In the presence of forces or an arbitrary dynamics, this operation because more complicated or ill-defined. The core problem with this definition is that it is not a property of the process itself, but on the existence of other processes (i.e. the process that undoes the changes or the process that flips the velocities, therefore changing the state). The requirement of time-invariance is too restrictive as it excludes all time dependent processes. Form invariance on an arbitrary set of equations, instead, can be broken by a suitably complicated change of variables, so it is only well-defined for a limited number of systems (e.g. particles under potential forces). The definition of reversibility we use is preferable because it works more generally and, when appropriate conditions apply, reduces to the other two which are better regarded as special cases. It is mathematically more elegant as it is the dual of determinism. It connects directly with information theory, as it states that the present has enough information to reconstruct the past, which gives a logical bridge to connect reversibility with information entropy and thermodynamics/statistical mechanics entropy.}

\fact{Determinism and reversibility imply Hamiltonian mechanics} {In the context of infinitesimal reducibility, determinism and reversibility means that all the particles and only the particle within a given initial state $\rho(s(t)) = \rho(s(t+\Delta t))$ will be mapped to a particular final state. That is, the densities will be preserved over time. Time evolution, then, is a canonical transformation in phase space. As time is continuous, we can find a function $H : \mathcal{S} \to \mathbb{R}$ which acts as a time generator, which we recognize as being the Hamiltonian. This is essentially Louisville's theorem applied in reverse.}

\fact{Invariant densities under time transformations imply extended phase-space} {So far, time has been treated as a parameter. If we introduce coordinate transformation that mix time, however, this is no longer possible. If $d\hat{q} = dq + vdt$ then the densities need to be defined on $dt$ as well. To do that, we have to introduce a conjugate variable $\omega$ of unit inverse time so that $dt d\omega$ is dimensionless and invariant. The invariant area element becomes $\sum_i dq^i dk_i - dt d\omega$ or $\sum_i dq^i dp_i - dt dE$ in units of $\hbar$. For surfaces at equal time $dt=0$ it reduces to the previous case. For surfaces that are not at equal time, since the temporal degree of freedom does cannot introduce new states, we need to correct the area element to account for the larger areas measured by the state variables that does not correspond to new states. If we set $q^\alpha = [t, q^i]$ and $p_\alpha = [-E, p_i]$ the area element becomes $dq^\alpha dp_\alpha$. These are elements of special relativity that come just by requiring invariance under time transformation. They do not come from invariance of the speed of light, they do not require a metric tensore. In other words, there is no way to have time transformations in Hamiltonian mechanics without requiring a four-momentum co-vector. Special relativity is not an independent choice: it is already partially baked into the two assumptions we took.}

\fact{Extended phase-space imply classical anti-particles} {The trajectories in the extended phase space are in terms of an affine parameter $s$ and not time: $q^\alpha(s)$. As the motion is still deterministic and reversible, though, time $t(s)$ must be monotonic with respect to the affine parameter. For each state, the direction of time can be either aligned or anti-aligned with the affine parameter. Because the relationship is monotonic, aligned states can only connect to other aligned states and anti-aligned states can only connect to other anti-aligned states. We call aligned states the particle states and the anti-aligned states the anti-particle states.}

\fact{Time evolution is better expressed by the Hamiltonian constraint} {The temporal degree of freedom introduces two variables but, since one is energy, there is an additional constraint since the value of the energy is linked to the other variables through the Hamiltonian. This link is better expressed by the Hamiltonian constraint $\mathcal{H} : \mathcal{\bar{S}} \to \mathbb{R}$ which is a function over the extended phase space, just like the Hamiltonian was a function over the standard phase space. The Hamiltonian constraint has two roles: it gives us the equation of motions in the same way of the standard Hamiltonian, and it constraints the motion over the surface $\mathcal{H} = 0$. To give context, the Hamiltonian constraint for a free particle is $\mathcal{H} = \frac{|p^\alpha|^2}{2m} - \frac{1}{2}mc^2$, which constraints the norm of the four-momentum to $mc$. The Hamiltonian constraint, unlike the standard Hamiltonian, is invariant under time transformation. It is the generator of the evolution in terms of an affine parameter.}

\fact{Hamiltonian constraint as operator on distributions} {If $\rho$ is the distribution for the whole system, then $\mathcal{H} \rho = 0$. In fact, the distribution can only be non-zero on the constrain, for which the Hamiltonian constraint is zero. So, for each point in the extended phase space, either the distribution is zero or the Hamiltonian constrain is zero. Note that this is the same form one has in quantum mechanics for the Klein-Gordon and Dirac equations $\mathcal{H} |\psi\rangle = 0$, which in fact can be understood as Hamiltonian constraints that fix the norm of the momentum four-vector and four-velocity.}

\fact{Determinism and reversibility is equivalent to conservation of information entropy} {Given a distribution, its information entropy is given by $I = - \int \rho \log \rho \, d\xi^n$. In general, information entropy is not invariant under change of variable: $- \int \rho(\hat{\xi}^b) \log \rho(\hat{\xi}^b) \, d\hat{\xi}^n = - \int \rho(\xi^a) \log \rho(\xi^a) \, d\xi^n - \int \rho(\xi^i) \log \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a}\right| \, d\xi^n$. Under a canonical transformation over phase space, though, the Jacobian determinant is unitary everyone therefore information entropy is invariant. That is, invariance of the density and invariance of its information entropy are one and the same. Requesting that densities are conserved in time is therefore equivalent to requesting that information entropy is conserved in time. This should be intuitive: the amount of information needed to identify an element within the initial or the final distribution is the same if the evolution is deterministic and reversible. Deterministic and reversible evolution, Hamiltonian evolution and conservation of information entropy, then, are one and the same.}

\fact{Conservation of information entropy leads to classical uncertainty principle} {As information entropy is conserved, a Hamiltonian process can only map a distribution to one that has the same information entropy. If we consider a single degree of freedom $(q, p)$ and fix the information entropy $I_0$, then the distribution that minimizes the product of the standard deviation is the Gaussian. The information entropy for the Gaussian is given by $I_G = \ln (2\pi e \sigma_q \sigma_p)$. During the evolution, we have:
$$ \sigma_q \sigma_p \geq \frac{e^{I_0}}{2\pi e} $$.}

@@

\fact{Note on notation} {We use $(q^\alpha, p_\alpha)$ for position and momentum and $(x^\alpha, u^\alpha)$ for position and velocity even though $q^\alpha = x^\alpha$. We do this because $\frac{\partial}{\partial q^\alpha} \neq \frac{\partial}{\partial x^\alpha}$ since one is taken at constant $p_\alpha$ and the other at constant $u^\alpha$. This is a tremendous source of confusion which is avoided by the use of the two symbols.}

\fact{Kinematic equivalence means having an invertible relationship between velocity and momentum} {If the state of the system $(q^\alpha, p_\alpha)$ is enough to determine the trajectory $x^\alpha(s)$, then it will be enough to determine the position and velocity $(x^\alpha, u^\alpha)$. These will be enough to reconstruct the state as they span a space of equal dimension. As $q^\alpha = x^\alpha$, then $u^\alpha = u^\alpha(q^\beta, p_\gamma)$ is invertible.}

\fact{Kinematic equivalence leads to Lagrangian mechanics} {Since $u^\alpha(q^\beta, p_\gamma)$, we can write $\mathcal{L} = \dot{q}^\alpha p_\alpha - \mathcal{H}$ as a function of $(x^\alpha, u^\alpha)$. That is the Lagrangian of the system. Note that this cannot be done if we do not have kinematic equivalence. For example, for a photon treated as a particle, we have $H = c |p|$. Using Hamilton's equations, we have $u^i = \frac{dx^i}{dt} = c \frac{p}{|p|}$ which is not invertible. While we can write the Lagrangian, this cannot be expressed in terms of position and velocity.}
